The Weakly Nonlinear Saturation of the Magnetorotational Instability

### 2014-12-17 paper/mri_paper.tex
hmri amplitude eqn 2
more paper background

For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent (\citei{Shakura:1973wg}). Since its discovery by \citeauthor{Chandrasekhar:1960wh} (\citeyear{Chandrasekhar:1960wh}) and subsequent rediscovery by \citeauthor{Balbus:1991vs} (\citeyear{Balbus:1991vs}), the MRI remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field destabilizes the disk gas, driving turbulence and angular momentum transport.

The MRI evolves via a period of linear growth, followed by nonlinear saturation. The nonlinear evolution of the MRI is a complex process, and an understanding of the mechanism by which the MRI saturates remains elusive. Most previous works on MRI saturation have studied the problem in simulation. Indeed, only a few works have tackled MRI saturation analytically. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. Another notable analytic work is that of \citeauthor{Pessah:2010ic} (\citeyear{Pessah:2010ic}), who presents an analysis of MRI saturation via secondary instabilities. This work suggests that MRI saturation can be caused by parasitic instabilities growing off of the primary MRI modes, in particular by Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime.

@@

@@

@@

>>>>

For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent (\citei{Shakura:1973wg}). Since its discovery by \citeauthor{Chandrasekhar:1960wh} (\citeyear{Chandrasekhar:1960wh}) and subsequent rediscovery by \citeauthor{Balbus:1991vs} (\citeyear{Balbus:1991vs}), the MRI remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability in its simplest geometry arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field destabilizes the disk gas, driving turbulence and angular momentum transport.

The MRI evolves via a period of linear growth, followed by nonlinear saturation. The nonlinear evolution of the MRI is a complex process, and an understanding of the mechanism by which the MRI saturates remains elusive. Several possible saturation avenues are available to the MRI. The instability could saturate via a reduction in the background shear (e.g. \citei{Umurhan:2007hs}), cutting off the supply of free energy on which the instability feeds. Similarly, the instability could pour its free energy into the magnetic field, and saturate via field generation (e.g. \citei{Ebrahimi:2009ey}). Saturation may also occur via ``parasite" instabilities, which feed off and destroy primary MRI modes (\citei{Goodman:1994ul}).

Most previous works on MRI saturation have studied the problem in simulation. Indeed, only a few works have tackled MRI saturation analytically. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. Another notable analytic work is that of \citeauthor{Pessah:2010ic} (\citeyear{Pessah:2010ic}), who presents an analysis of MRI saturation via secondary instabilities. This work suggests that MRI saturation can be caused by parasitic instabilities growing off of the primary MRI modes, in particular by Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime.

@@

\section{Simulations}

@@

The second amplitude equation (\ref{ampeq2}) becomes

\beq\label{ampeq2_hmri}
\partial_T \beta = - m \partial_Z^2 \beta + \frac{1}{\reym} \partial_Z^2 \beta \, - \, s \xi 2 \left|\alpha\right|\partial_Z \alpha \, - \, \xi\partial_Z^2 \beta,
\eeq

\beq
s = \left<\mathbb{U}_{11} \cdot \left(\mathcal{H}\mathbb{V}_{20}\right)^*\right>
\eeq

\beq
f = \left< \mathbb{U}_{11} \cdot \left(\mathcal{H}\mathbb{U}_{20}\right)^*\right>.
\eeq

If $s = 0$, as it is for the standard MRI, the two amplitude equations will again decouple.

@@

\bibitem[{Ebrahimi et al.(2009)}]{Ebrahimi:2009ey}
Ebrahimi, F., Prager, S.C., Schnack, D.D
